From Reversible Programs to \\ Univalent Universes and Back

### 2021-07-04 Pi+/paper/ufin.tex
Revise UFin section and fit into page budget

The groupoid $\BFin$ that we presented, is strict, that is, two equal permutations are extensionally identified, without
an explicit witness for the equality. Our first step is to weaken this groupoid $\BFin$ into a weak 2-category. In a
weak category, equational axioms of the category are replaced by higher invertible cells. For example, there are higher
invertible cells for the left and right units of composition and associativity. The higher cells have their own
coherences as well, the 2-cells can be composed with 1-cells by whiskering, and so on.

HoTT provides a rich internal language for working with weak higher groupoids, using the ``Types are weak
$\infty$-groupoids'' correspondence. We will make this precise by observing that a 1-groupoid in HoTT gives a locally
strict $(2,0)$-category, using the identity type for the higher cells. The groupoid $\BFin$ can be written in HoTT as a
type $\UFin$, which is a 1-groupoid.

Secondly, we would like to like to give a syntactic presentation for this groupoid, by describing exactly how the
1-cells and 2-cells are generated. Note that, in $\BFin$, the set of bijections on a finite set of cardinality $n$ has a
group structure, the identity function is the neutral element, group multiplication is given by function composition,
and the group inverse is given by inverting the bijection. This is the automorphism group of the finite set, which is
also known to be the finite symmetric group $\Sn[n]$.

Groups can be defined by listing the carrier set and the operations of the group, or they can be described using a group
presentation, which gives the generators and relations of a group. The last step of our technical development is to use
a group presentation for $\Sn$ to give a presentation for the higher cells of the groupoid $\UFin$. Using this
characterisation, we will obtain our completeness result.

% In the previous section, we have picked a particular encoding of finite types: as sequences of elements in some
% canonical order.

% This is not a ``free'' presentation. Some properties of the encoding might pollute our semantics. We want to make that
% any theorems / properties of the semantics we prove apply to the source language (under various interpretations).

% This motivates us to use a different semantic representation: a univalent universe of finite types which contains the
% finite types and nothing but the finite types and is defined such as alternative representations of finite types are
% identified. This guarantees that we can get a proper, sound and complete, semantics for our language.

% \begin{center}
% \begin{tikzpicture}[scale=1,every node/.style={scale=1}]
% \usetikzlibrary{arrows}
% \usetikzlibrary{shapes}
% \node[ultra thick, draw=blue, ellipse, minimum width=300pt, minimum height=300pt, align=center] {};
% \node[ultra thick, draw=blue, ellipse, minimum width=70pt, minimum height=40pt, align=center] (a) {$\{ 1 , 2 \} \uplus \{ 3 \}$};
% \node[ultra thick, draw=blue, ellipse, minimum width=70pt, minimum height=40pt, align=center] (b) at ([shift=({70pt,-70pt})]a) {$\{ 1 \} \uplus \{ 2, 3 \}$};
% \node[ultra thick, draw=blue, ellipse, minimum width=70pt, minimum height=40pt, align=center] (c) at
% ([shift=({-70pt,-70pt})]a) {$\{ 10 \} \uplus \{ 20, 30 \}$};
% \node (d) at ([shift=({50pt,-10pt})]a) {$\cdots$};
% \draw[thick,-] (a) to node[draw=none,right] {$\equiv$} (b);
% \draw[thick,-] (b) to node[draw=none,below] {$\equiv$} (c);
% \draw[thick,-] (a) to node[draw=none,left] {$\equiv$} (c);
% \end{tikzpicture}
% \end{center}

@@

Given an operational model of how a type A reduces to a type B
Given a denotational model of which types A and B are equivalent
Want soundness and completeness results
Guarantee that the two notions of equality are themselves “equivalent”
Generalized “identity of indiscernibles”

Previous semantics adequate for reasoning about evaluation and for checking the equivalence of two circuits.

Want NBE, want rewriting system of program equivalence, what else? So we need to move to proof-relevant higher-category
groupoid semantics instead of set-theoretic.

Now the question is, what does it mean for a compiler?

My opinion is that it is still the same, what rewrites to apply depends
on the choices made by the compiler writer, but we can think about that.

%%%
\subsection{To Put Somewhere or Delete}

This approach is sufficient to prove the semantics forms a 1-category but ignores the rich structure at the next
level~\cite{DBLP:conf/esop/CaretteS16}.
As explained in the previous section, a $\Pi$-type $A$ has $\sizet{A}$-elements and for all combinators $c : A \iso B$
we have that $\sizet{A} = \sizet{B}$. Hence, the denotation $\denot{A}$ of a type $A$ with $n$-elements can be the finite
set $\Fin[n] = \{ 0, 1, \cdots, n-1\}$; the denotation of a value $v : A$ such that $\sizet{A}=n$ will be an index in
the range $[0,n-1]$, and the denotation $\denot{c}$ of a combinator $c : A \iso B$ such that
$\sizet{A} = \sizet{B} = n$ will be a function from $\Fin[n]$ to $\Fin[n]$ that permutes the elements. Thus, all types
with 3 elements will denote $\Fin[3]$ and combinators between them will denote permutations on $\Fin[3]$, e.g.:
\[\begin{array}{rcl}
    \denot{\onet + (\onet + \onet)} &=& \{ 0,1,2 \} \\
    \denot{(\onet + \onet) + \onet} &=& \{ 0,1,2 \} \\
    \\
    \denot{\assoclp : \onet + (\onet + \onet) \iso (\onet + \onet) + \onet} &=& (0~1~2) \\
    \denot{\swapp : \onet + (\onet + \onet) \iso (\onet + \onet) + \onet} &=& (2~0~1)
\end{array}\]
where we have used the one-line notation for permutations with $(a~b~c)$ representing the
permutation that maps 0 to $a$, 1 to $b$, and 2 to $c$. To make the denotation of values precise, we compute a canonical
enumeration of the elements of each type:
\[\begin{array}{rcl}
    \mathit{enum}(\zerot) &=& [~] \\
    \mathit{enum}(\onet) &=& [ ~\Acon{tt}~ ] \\
    \mathit{enum}(A + B) &=& \mathit{map}~\Acon{inj₁}~\mathit{enum}(A) ~\textsf{+\!+}~ \mathit{map}~\Acon{inj₂}~\mathit{enum}(B) \\
    \mathit{enum}(A \times B) &=& \mathit{concat}~(\mathit{map}~(\lambda v.\mathit{map}~(\lambda w. (v,w))~\mathit{enum}(B))~\mathit{enum}(A))
\end{array}\]
\noindent The specification uses a Haskell-like notation for sequences with $\mathit{map}$ as the operation that applies
a function to each element of a sequence, \textsf{+\!+} as the binary append operation, and $\mathit{concat}$ the
operation that appends all the subsequences in a sequence.

Using the definition, we have:
\[\begin{array}{rcl}
    \mathit{enum}(\onet + (\onet + \onet)) &=& [ \inlv{\Acon{tt}},~\inrv{(\inlv{\Acon{tt}})},~\inrv{(\inrv{\Acon{tt}})} ] \\
    \mathit{enum}((\onet + \onet) + \onet) &=& [ \inlv{(\inlv{\Acon{tt}})},~\inlv{(\inrv{\Acon{tt}})},~\inrv{\Acon{tt}} ]
\end{array}\]
Thus, as shown in the diagrams below, $\assoclp~(\inlv{\Acon{tt}})$ applies the permutation $(0~1~2)$ to the index of
$\inlv{\Acon{tt}}$ which is 0 and produces index 0 in the $(\onet + \onet) + \onet$ type corresponding to value
$\inlv{(\inlv{\Acon{tt}})}$. Similarly, $\swapp~(\inlv{\Acon{tt}})$ applies the permutation $(2~0~1)$ to the index of
$\inlv{\Acon{tt}}$ which is 0 and produces index 2 in the $(\onet + \onet) + \onet)$ type corresponding to value
$\inrv{\Acon{tt}}$.

@@

\[\begin{array}{rcll}
\sem{A} &=& (\mathsf{Fin}~n, [ n , \mathsf{refl} ]) & \mbox{where}~\sizet{A} = n
\end{array}\]

@@

%%%
\subsection{Univalent Subuniverses}
\label{sec:univalent}

In order to define the semantics, we seek a mathematical structure satisfying the following properties: (i) it contains structures corresponding to all the finite types and nothing but the finite types, and (ii) it is robust enough to ensure that equivalent encodings of finite types are identified.

In this section, we introduce some basic concepts and notation that we use in Homotopy Type Theory. Then, we define
univalent subuniverses and discuss some specific examples.

\subsection{The Type Theory}~\label{subsec:type-theory}

We work in Homotopy Type Theory (HoTT), and use the language of the HoTT
book~\cite{univalentfoundationsprogramHomotopyTypeTheory2013}, that is, we use intensional Martin-L\"{o}f Type Theory,
with a univalent universe $\UU$, and we use Higher Inductive Types (HITs) for propositional truncations and set
quotients. All arguments will hold in a cubical type theory as well, such
as~\cite*{cohenCubicalTypeTheory2018,angiuliComputationalSemanticsCartesianCubical2019}. We recall a few basic facts to
familiarise the reader with the notation we use. For more details about basics of HoTT, we refer the reader to the HoTT
book~\cite{univalentfoundationsprogramHomotopyTypeTheory2013}.

Our denotational semantics is given using groupoids, and HoTT provides a rich internal language for describing (weak
$\infty$-)groupoids. By formalising our results using general tools and principles from HoTT, we're able to use a proof
assistant to computer-check our proofs. Our results are formalised in the Agda proof assistant using the HoTT-Agda
library, though it could also be formalised using Cubical Agda~\cite{vezzosiCubicalAgdaDependently2019}. \review{Using
  the formalisation of the semantics, we are also able to extract a computable procedure for performing normalisation of
  circuits.}

@@

\vc{I'm explaining everything starting from the categorical language. We do care more about the groupoid structure of
  types, and how to encode groupoids using types, since that's what we use in~\cref{sec:finite}.}

In HoTT, the identity type or the equality type is the type of paths or equalities between two terms of the same type.
Given two terms $x:A$ and $y:A$, we write $x \id_{A} y$, or simply $x \id y$, for the equality type between them. In
book HoTT, the identity type is generated by reflexivity $\refl_{x} : x \id_{A} x$, and the eliminator for the identity
type is given by path induction or the $J$-rule. The identity type equips each type with the structure of a weak
$\infty$-groupoid, or a homotopy type. For example, given an arbitrary type (or groupoid) $A$, we list some groupoid
laws that are provable using path induction.

@@

Functions between types are functors between groupoids. Given a function $f : A \to B$, the functorial action
$\term{ap}_{f}(\blank)$ is given by path induction.

\[
  \term{ap}_{f} : \dfun{x,y:A}{x \id_{A} y \to f(x) \id_{B} f(y)}
\]

Type families are functions from a type to the universe, such as $A \to \UU$, which is an $A$-indexed family of
groupoids. The $\term{transport}$ operation gives the functorial action of paths in the indexing type, which is also
defined by path induction. If $P : A \to \UU$ is a type family, then for a path $x \id_{A} y$, we have

\[
  \transport{P} : \dfun{x,y:A}{x \id_{A} y \to P(x) \to P(y)}
\]

A type family can also be seen as a fibration, from the topological viewpoint. For a type family $P : A \to \UU$ and a
point $x : A$, the type $P(x)$ gives the fiber over $x$. The type $\dsum{x:A}{P(x)}$ is the collection of all the fibers
and is called the total space of $P$. The first projection ${\pi_1 : \dsum{x:A}{P(x)} \to A}$ from the total space to
the base space $A$ is then a fibration. The lifting operation lifts paths in the base space to paths in the total space.
Given a path $p : x \id_{A} y$ in the base space, and $u : P(x)$ a point in the fiber over $x$, we have

\[
  \term{lift}(u,p) : (x , u) \id_{\dsum{x:A}{P(x)}} (y , \tr{p}{u})
\]

where $\tr{p}{u}$ is shorthand for $\transport{P,p}(u)$.

\begin{center}
  \begin{tikzpicture}[yscale=.5,xscale=2]
    \draw (0,0) arc (-90:170:8ex) node[anchor=south east] {$A$} arc (170:270:8ex);
    \draw (0,6) arc (-90:170:10ex) node[anchor=south east] {$\dsum{x:A}{P(x)}$} arc (170:270:10ex);
    \draw[->] (0,5.8) -- node[auto] {$\fst$} (0,3.2);
    \node[circle,fill,inner sep=1pt,label=left:{$x$}] (b1) at (-.5,1.4) {};
    \node[circle,fill,inner sep=1pt,label=right:{$y$}] (b2) at (.5,1.4) {};
    \draw[decorate,decoration={snake,amplitude=1}] (b1) -- node[auto,swap] {$p$} (b2);
    \node[circle,fill,inner sep=1pt,label=left:{$\pair{x,u}$}] (b1) at (-.5,7.2) {};
    \node[circle,fill,inner sep=1pt,label=right:{$\pair{y,\tr{p}{u}}$}] (b2) at (.5,7.2) {};
    \draw[decorate,decoration={snake,amplitude=1}] (b1) -- node[auto] {$\term{lift}(u,p)$} (b2);
  \end{tikzpicture}
\end{center}

Further, using the groupoid structure of $A$, we can show that $\term{transport}$ lifts paths to equivalences, we define

@@

\begin{definition}[Univalent Fibration]
  $P$ is a univalent type family (or simply a univalent fibration) if $\tptEqv{P}$ is an equivalence.
\end{definition}

@@

Voevodsky's \emph{univalence} principle characterises paths in the universe. It says that equivalent types are equal, or
the following function is an equivalence.

@@

@@

\todo{Explain homotopy types, $\hProp$, $\hSet$, etc.}

\vc{These are placeholder definitions, need informal explanations and references to the book.}

\begin{definition}[Propositional Truncation]
  Given a type $A$, the propositional truncation $\Trunc[-1]{A}$, or simply $\Trunc{A}$, is a higher inductive type
  generated by the following constructors,
  \begin{itemize}
    \item an inclusion function $\trunc{\blank} : A \to \Trunc{A}$,
    \item for each $x, y : \Trunc{A}$, a path $\term{trunc}(x,y) : x \id_{\Trunc{A}} y$,
  \end{itemize}
  such that, given any type $B$ with
  \begin{itemize}
    \item a function $g : A \to B$,
    \item for each $x, y : B$, a path $\term{trunc*}(x,y) : x \id_{B} y$,
  \end{itemize}
  there is a unique function $f : \Trunc{A} \to B$ such that,
  \begin{itemize}
    \item $f(\trunc{a}) \equiv g(a)$
    \item for each $x, y : \Trunc{A}$, $\ap{f}{\term{trunc}(x,y)} \id_{B} \term{trunc*}(f(x),f(y))$.
  \end{itemize}
\end{definition}

\begin{definition}[Set Quotient]
  Given a type $A$ which is an $\hSet$, and a relation $R : A \to A \to \hProp$, the set-quotient $\quot{A}{R}$ is the
  higher inductive type generated by
  \begin{itemize}
    \item an inclusion function $q : A \to \quot{A}{R}$,
    \item for each $x, y : A$ such that $R(x,y)$, a path $q(x) \id_{\quot{A}{R}} q(y)$,
    \item a set truncation, for each $x, y : \quot{A}{R}$ and $r, s : x \id_{\quot{A}{R}} y$, we have $r \id s$,
  \end{itemize}
  with \review{an appropriate induction principle.}
\end{definition}

We recall that quotients in HoTT are \emph{effective}, that is, if $R$ is an equivalence relation, we have
$R(x,y) \eqv (q(x) \id_{\quot{A}{R}} q(y))$.

@@

\todo{Explain motivation. Starting from a univalent universe which classifies all types, we want to define a subuniverse
  which classifies only certain types, for example, types that satisfy some property that we want.}

\begin{definition}[Universe]
  A universe \`{a} la Tarski is given by the following pieces of data,
  \begin{itemize}
    \item a code $U : \UU$,
    \item a decoding function $\El : U \to \UU$.
  \end{itemize}
  If $\El$ is a univalent fibration, $U$ is a univalent universe.
\end{definition}

\begin{definition}[Subuniverse]
  A subtype is a type family $P : \UU \to \UU$ whose fibers are prop-valued, that is, $\forall x, \isProp{P(x)}$. A
  subuniverse generated by a subtype has $U \defeq \dsum{X:\UU}{P(X)}$ and $\El \defeq \fst$.
\end{definition}

@@

\begin{proof}
  Suppose $(U, \El) \defeq (\dsum{X:\UU}{P(X)}, \fst)$ is a subuniverse generated by a subtype $P : \UU \to \UU$. For
  any $X, Y : \UU$ such that $\phi : P(X)$ and $\psi : P(Y)$, we want to show that
  $\tptEqv{\fst} : (X,\phi) \id (Y,\psi) \to X \eqv Y$ is an equivalence. We construct
  $X \eqv Y \to (X,\phi) \id (Y,\psi)$ by $\ua$ and using the fact that $P(\blank)$ is a proposition. That it is an
  inverse follows by calculation using the appropriate computation rules.
\end{proof}

\begin{example}[$\BAut$]
  The type of self-equivalences on a type $T$ is defined as $\Aut[T] \defeq T \eqv T$, which is an $\infty$-group. The
  subuniverse of types that are merely equal to $T$ is given by $\BAut[T] \defeq \Sub{T}$. This turns out to be the
  classifying space of $\Aut[T]$, as justified by~\cref{lem:loop-deloop}. We write $T_{o} \defeq (T, \trunc{\refl_{T}})$
  for the image of the inclusion of $T$.
\end{example}

\begin{example}
  The univalent subuniverse of types merely equal to $\Bool$ is given by $\BAut[\Bool]$. This can be used to give the
  denotational semantics of a 2-bit reversible language~\cite{caretteReversibleProgramsUnivalent2018}.
\end{example}

\begin{definition}[$\Fin$]
  The type family $\Fin : \Nat \to \UU$ is the type of finite sets indexed by cardinality. It is defined equivalently in
  two different ways,
  \begin{gather*}
    \begin{aligned}
      \Fin[n] & \defeq \dsum{k:\Nat}{k < n}
    \end{aligned}
    \begin{aligned}
       & \Fin[0] \defeq \bot                      \\
       & \Fin[\suc[n]] \defeq \top \sqcup \Fin[n]
    \end{aligned}
  \end{gather*}
  Note that $\Fin[n]$ is a $\hSet$, and we use the definitions interchangeably.
\end{definition}

\begin{example}
  For any $n : \Nat$, we define the univalent subuniverse of types merely equal to $\Fin[n]$ as $\BAut[\Fin[n]]$. This
  is the universe of finite sets with cardinality equal to $n$. We write $F_{n} \defeq (\Fin[n], \trunc{\refl})$ for the
  image of the inclusion of $\Fin[n]$.
\end{example}

@@

\todo{rig}

We describe the symmetric monoidal structure of the groupoid $\UFin$.

First, we observe a few equivalences.

>>>>

The groupoid $\BFin$ that we presented is strict, that is, two equal permutations are extensionally identified, without
an explicit witness for the equality. Our first step is to weaken this groupoid $\BFin$ into a weak 2-category. In a
weak category, equational axioms of the category are replaced by higher invertible cells. For example, there are higher
invertible cells for the left and right units of composition, and associativity. The higher cells come with their own
coherence data as well, such as vertical composition, and horizontal composition, or whiskering.\todo{define 2-category
  in appendix.}

HoTT provides a rich internal language for working with weak higher groupoids, using the ``Types are weak
$\infty$-groupoids'' correspondence. Using the identity type for morphisms, a 1-groupoid in HoTT has points for 0-cells,
1-paths for 1-cells, and 2-paths for 2-cells, but there is at most one 2-cell between compatible 1-cells.~\footnote{This
  is a locally-strict $(2,0)$-category, since every cell above 0 is invertible, and the hom-categories are posets, that
  is, truth-value enriched.} The groupoid $\BFin$ can be written in HoTT as a type $\UFin$, which is a 1-groupoid.

\begin{toappendix}
  We give an example of the groupoid structure on a 3-element set.

@@

\end{toappendix}

\begin{toappendix}

  \subsection{To Put Somewhere or Delete}

  This approach is sufficient to prove the semantics forms a 1-category but ignores the rich structure at the next
  level~\cite{DBLP:conf/esop/CaretteS16}.
  As explained in the previous section, a $\Pi$-type $A$ has $\sizet{A}$-elements and for all combinators $c : A \iso B$
  we have that $\sizet{A} = \sizet{B}$. Hence, the denotation $\denot{A}$ of a type $A$ with $n$-elements can be the finite
  set $\Fin[n] = \{ 0, 1, \cdots, n-1\}$; the denotation of a value $v : A$ such that $\sizet{A}=n$ will be an index in
  the range $[0,n-1]$, and the denotation $\denot{c}$ of a combinator $c : A \iso B$ such that
  $\sizet{A} = \sizet{B} = n$ will be a function from $\Fin[n]$ to $\Fin[n]$ that permutes the elements. Thus, all types
  with 3 elements will denote $\Fin[3]$ and combinators between them will denote permutations on $\Fin[3]$, e.g.:
  \[\begin{array}{rcl}
      \denot{\onet + (\onet + \onet)}                                         & = & \{ 0,1,2 \} \\
      \denot{(\onet + \onet) + \onet}                                         & = & \{ 0,1,2 \} \\
      \\
      \denot{\assoclp : \onet + (\onet + \onet) \iso (\onet + \onet) + \onet} & = & (0~1~2)     \\
      \denot{\swapp : \onet + (\onet + \onet) \iso (\onet + \onet) + \onet}   & = & (2~0~1)
    \end{array}\]
  where we have used the one-line notation for permutations with $(a~b~c)$ representing the
  permutation that maps 0 to $a$, 1 to $b$, and 2 to $c$. To make the denotation of values precise, we compute a canonical
  enumeration of the elements of each type:
  \[\begin{array}{rcl}
      \mathit{enum}(\zerot)     & = & [~]                                                                                                          \\
      \mathit{enum}(\onet)      & = & [ ~\Acon{tt}~ ]                                                                                              \\
      \mathit{enum}(A + B)      & = & \mathit{map}~\Acon{inj₁}~\mathit{enum}(A) ~\textsf{+\!+}~ \mathit{map}~\Acon{inj₂}~\mathit{enum}(B)          \\
      \mathit{enum}(A \times B) & = & \mathit{concat}~(\mathit{map}~(\lambda v.\mathit{map}~(\lambda w. (v,w))~\mathit{enum}(B))~\mathit{enum}(A))
    \end{array}\]
  \noindent The specification uses a Haskell-like notation for sequences with $\mathit{map}$ as the operation that applies
  a function to each element of a sequence, \textsf{+\!+} as the binary append operation, and $\mathit{concat}$ the
  operation that appends all the subsequences in a sequence.

  Using the definition, we have:
  \[\begin{array}{rcl}
      \mathit{enum}(\onet + (\onet + \onet)) & = & [ \inlv{\Acon{tt}},~\inrv{(\inlv{\Acon{tt}})},~\inrv{(\inrv{\Acon{tt}})} ] \\
      \mathit{enum}((\onet + \onet) + \onet) & = & [ \inlv{(\inlv{\Acon{tt}})},~\inlv{(\inrv{\Acon{tt}})},~\inrv{\Acon{tt}} ]
    \end{array}\]
  Thus, as shown in the diagrams below, $\assoclp~(\inlv{\Acon{tt}})$ applies the permutation $(0~1~2)$ to the index of
  $\inlv{\Acon{tt}}$ which is 0 and produces index 0 in the $(\onet + \onet) + \onet$ type corresponding to value
  $\inlv{(\inlv{\Acon{tt}})}$. Similarly, $\swapp~(\inlv{\Acon{tt}})$ applies the permutation $(2~0~1)$ to the index of
  $\inlv{\Acon{tt}}$ which is 0 and produces index 2 in the $(\onet + \onet) + \onet)$ type corresponding to value
  $\inrv{\Acon{tt}}$.

@@

  \[\begin{array}{rcll}
      \sem{A} & = & (\mathsf{Fin}~n, [ n , \mathsf{refl} ]) & \mbox{where}~\sizet{A} = n
    \end{array}\]

@@

\end{toappendix}

%%%
\subsection{The Type Theory}~\label{subsec:type-theory}
% \subsection{Univalent Subuniverses}
% \label{sec:univalent}

% In order to define the weak groupoid we're after, we seek a mathematical structure satisfying the following properties:
% (i) it contains structures corresponding to all the finite types and nothing but the finite types, and (ii) it is robust
% enough to ensure that equivalent encodings of finite types are identified.

To define the weak groupoid we're after, we first introduce some basic concepts and notation from Homotopy Type Theory
that we use.~\footnote{We use the language of the HoTT book~\cite{univalentfoundationsprogramHomotopyTypeTheory2013},
that is, we use intensional Martin-L\"{o}f Type Theory, with a (univalent) universe $\UU$, and a few Higher Inductive
Types (HITs) for propositional truncation and set quotients. All arguments will hold in a Cubical Type Theory as well,
such
as~\cite*{cohenCubicalTypeTheory2018,angiuliComputationalSemanticsCartesianCubical2019,vezzosiCubicalAgdaDependently2019}.}
For more details about HoTT, we refer the reader to the HoTT
book~\cite{univalentfoundationsprogramHomotopyTypeTheory2013}. By formalising our results in HoTT, we're also able to
use a proof assistant to computer-check our proofs, and our results are formalised in the Agda proof assistant using the
HoTT-Agda library. Using the formalisation of our semantics, we are able to extract a computable procedure for
performing NbE and synthesis of reversible circuits, as motivated in~\cref{sec:introduction}.

@@

Given two terms $x:A$ and $y:A$, we write $x \id_{A} y$, or simply $x \id y$, for the identity type, which is the type
of equalities or identifications between them. The identity type is generated by reflexivity $\refl_{x} : x \id_{A} x$,
and the eliminator for the identity type is given by path induction or the $J$-rule (\cref{def:path-induction}). This
construction can be iterated, giving the identity type between two terms of an identity type, repeating ad infinitum.
Using the iterated identity type for morphisms, each type is equipped with the structure of a weak $\infty$-groupoid,
where each morphism satisfies groupoid laws only upto a higher one. Given an arbitrary type (or groupoid) $A$, we list
some laws that are provable using path induction.

\begin{toappendix}
  \begin{definition}[Path Induction]
    \label{def:path-induction}
    Given a type family $C : \dfun{x,y:A}{(x \id_A y)} \to \UU$, and a function $c : \dfun{x:A}{C(x,x,\refl_x)}$, there is
    a function $f : \dfun{x,y:A}{\dfun{p:x \id_A y}{C(x,y,p)}}$ such that $f(x,x,\refl_x) \defeq c(x)$.
  \end{definition}
\end{toappendix}

@@

Functions between types are functors between groupoids. Given a function $f : A \to B$, there is a functorial action on
the paths given by $\term{ap}$. Type families, that is, types indexed by terms, are simply functions from a type to the
universe, such as $A \to \UU$. For a type family $P : A \to \UU$ and a point $x : A$, the type $P(x)$ is called the
fiber over $x$.  The $\term{transport}$ operation, named $\term{tr}$, lifts paths in the indexing type to functions
between fibers.
%which is an $A$-indexed family of groupoids.

\begin{gather*}
  \term{ap}_{f} : \dfun{x,y:A}{x \id_{A} y \to f(x) \id_{B} f(y)}
  \qquad
  \term{tr}_{P} : \dfun{x,y:A}{x \id_{A} y \to P(x) \to P(y)}
\end{gather*}

The type $\dsum{x:A}{P(x)}$ is the collection of all the fibers and is called the total space of $P$. The first
projection ${\pi_1 : \dsum{x:A}{P(x)} \to A}$ from the total space to the base space $A$ has the structure of a
fibration, that is, there is a lifting operation (\cref{fig:lift}) which lifts paths in the base space to paths in the
total space. Given a path $p : x \id_{A} y$ in the base space, and $u : P(x)$ a point in the fiber over $x$, we have

\[
  \term{lift}(u,p) : (x , u) \id_{\dsum{x:A}{P(x)}} (y , \transport{P}{p}{u})
\]

% where $\tr{p}{u}$ is shorthand for $$.

\begin{figure}
  \begin{center}
    \begin{tikzpicture}[yscale=.5,xscale=2]
      \draw (0,0) arc (-90:170:8ex) node[anchor=south east] {$A$} arc (170:270:8ex);
      \draw (0,6) arc (-90:170:10ex) node[anchor=south east] {$\dsum{x:A}{P(x)}$} arc (170:270:10ex);
      \draw[->] (0,5.8) -- node[auto] {$\fst$} (0,3.2);
      \node[circle,fill,inner sep=1pt,label=left:{$x$}] (b1) at (-.5,1.4) {};
      \node[circle,fill,inner sep=1pt,label=right:{$y$}] (b2) at (.5,1.4) {};
      \draw[decorate,decoration={snake,amplitude=1}] (b1) -- node[auto,swap] {$p$} (b2);
      \node[circle,fill,inner sep=1pt,label=left:{$\pair{x,u}$}] (b1) at (-.5,7.2) {};
      \node[circle,fill,inner sep=1pt,label=right:{$\pair{y,\transport{P}{p}{u}}$}] (b2) at (.5,7.2) {};
      \draw[decorate,decoration={snake,amplitude=1}] (b1) -- node[auto] {$\term{lift}(u,p)$} (b2);
    \end{tikzpicture}
  \end{center}
  \caption{Lifting operation in $P$}
  \label{fig:lift}
\end{figure}

Further, using the groupoid structure of $A$, we can show that $\term{transport}$ lifts paths to equivalences of fibers.

@@

\begin{definition}[Univalent Fibration]
  $P$ is a univalent type family (or, $\fst : {\dsum{x:A}{P(x)}} \to A$ is a univalent fibration) if $\tptEqv{P}$ is an
  equivalence.
\end{definition}

@@

Voevodsky's \emph{univalence} principle characterises paths in the universe. It says that equivalent types are equal, or,
the following function is an equivalence.

@@

\subsubsection{Homotopy Types}

A type is \emph{contractible} (-2-type) if it has a unique element, that is, there is a center of contraction and every
other point is equal to it. A type is a \emph{proposition} (-1-type) if its equality types are contractible, that is, it
has at most one inhabitant. Iterating this, we can define sets or 0-types (whose equality types are propositions) and
1-groupoids or 1-types (whose equality types are sets), and similarly, higher homotopy $n$-types.

\begin{gather*}
  \begin{aligned}
    \isContr{A} & \defeq \dsum{x:A}{\dfun{y:A}{y \id x}} \\
    \isProp{A}  & \defeq \dfun{x,y:A}{\isContr{x \id y}}
  \end{aligned}
  \qquad
  \begin{aligned}
    \isSet{A} & \defeq \dfun{x,y:A}{\isProp{x \id y}} \\
    \isGpd{A} & \defeq \dfun{x,y:A}{\isSet{x \id y}}
  \end{aligned}
\end{gather*}

@@

Higher Inductive Types generalise Inductive Types, by allowing path constructors besides point constructors. While point
constructors generate the elements of the type, path constructors generate equalities between points in the type. We
describe a few basic HITs that we use.

Given a type $A$, the propositional truncation $\Trunc[-1]{A}$, squashes the elements of $A$ turning it into a
proposition. It is given by a HIT with a point constructor $\trunc{\blank} : A \to \Trunc{A}$, and a path constructor
$\term{trunc}(x,y) : x \id_{\Trunc{A}} y$, which equates every pair of points in the truncation
(see~\cref{def:prop-trunc}).

\begin{toappendix}
  \begin{definition}[Propositional Truncation]
    \label{def:prop-trunc}
    Given a type $A$, the propositional truncation $\Trunc[-1]{A}$, or simply $\Trunc{A}$, is a higher inductive type
    generated by the following constructors,
    \begin{itemize}
      \item an inclusion function $\trunc{\blank} : A \to \Trunc{A}$,
      \item for each $x, y : \Trunc{A}$, a path $\term{trunc}(x,y) : x \id_{\Trunc{A}} y$,
    \end{itemize}
    such that, given any type $B$ with
    \begin{itemize}
      \item a function $g : A \to B$,
      \item for each $x, y : B$, a path $\term{trunc*}(x,y) : x \id_{B} y$,
    \end{itemize}
    there is a unique function $f : \Trunc{A} \to B$ such that,
    \begin{itemize}
      \item $f(\trunc{a}) \equiv g(a)$
      \item for each $x, y : \Trunc{A}$, $\ap{f}{\term{trunc}(x,y)} \id_{B} \term{trunc*}(f(x),f(y))$.
    \end{itemize}
  \end{definition}
\end{toappendix}

Another HIT that we use is the set-quotient $\quot{A}{R}$ which takes an $\hSet$ $A$ and a relation $R : A \to A \to
  \UU$. It has an inclusion of points $q : A \to \quot{A}{R}$, and adds paths between related pairs of elements
$\quotrel : R(x,y) \to q(x) \id_{\quot{A}{R}} q(y)$ (see~\cref{def:set-quot}).  We recall that the quotient is
\emph{effective} if $R$ is a prop-valued equivalence relation, that is, $R(x,y)$ holds iff $(q(x) \id_{\quot{A}{R}}
  q(y))$.

\begin{toappendix}
  \begin{definition}[Set Quotient]
    \label{def:set-quot}
    Given a type $A$ which is an $\hSet$, and a relation $R : A \to A \to \hProp$, the set-quotient $\quot{A}{R}$ is the
    higher inductive type generated by
    \begin{itemize}
      \item an inclusion function $q : A \to \quot{A}{R}$,
      \item for each $x, y : A$ such that $R(x,y)$, a path $q(x) \id_{\quot{A}{R}} q(y)$,
      \item a set truncation, for each $x, y : \quot{A}{R}$ and $r, s : x \id_{\quot{A}{R}} y$, we have $r \id s$,
    \end{itemize}
    with \review{an appropriate induction principle.}
  \end{definition}
\end{toappendix}

@@

Starting from a univalent universe which classifies all types, we want to define a subuniverse which classifies only
certain types, for example, types that satisfy some desired property. We use a prop-valued type family, that is, a
predicate, which picks out only those types, and collect them into a univalent subuniverse. Being univalent ensures that
the equality type of the ambient universe is reflected in the subuniverse.

\begin{definition}[Universe]
  A universe \`{a} la Tarski is given by the following pieces of data,
  \begin{itemize}
    \item a code $U : \UU$,
    \item a decoding type family $\El : U \to \UU$.
  \end{itemize}
  If $\El$ is univalent, $U$ is a univalent universe.
\end{definition}

\begin{definition}[Subuniverse]
  A subtype is a type family $P : \UU \to \UU$ whose fibers are prop-valued, that is, $P(X)$ is a proposition for every
  $X$. A subuniverse generated by a subtype has $U \defeq \dsum{X:\UU}{P(X)}$ and $\El \defeq \fst$.
\end{definition}

@@

\begin{appendixproof}
  Suppose $(U, \El) \defeq (\dsum{X:\UU}{P(X)}, \fst)$ is a subuniverse generated by a subtype $P : \UU \to \UU$. For
  any $X, Y : \UU$ such that $\phi : P(X)$ and $\psi : P(Y)$, we want to show that
  $\tptEqv{\fst} : (X,\phi) \id (Y,\psi) \to X \eqv Y$ is an equivalence. We construct
  $X \eqv Y \to (X,\phi) \id (Y,\psi)$ by $\ua$ and using the fact that $P(\blank)$ is a proposition. That it is an
  inverse follows by calculation using the appropriate computation rules.
\end{appendixproof}

The types we're interested in are the finite types. In constructive mathematics, the notion of finiteness is subtle. We
define a type to be (Bishop-) finite if it is equal to some finite set, and we setup the predicate $\isFin$ to do so.

\begin{definition}[$\Fin$]
  The type family $\Fin : \Nat \to \UU$ is the type of finite sets indexed by their cardinality. It is defined
  equivalently in two different ways,
  \begin{gather*}
    \begin{aligned}
      \Fin[n] & \defeq \dsum{k:\Nat}{k < n}
    \end{aligned}
    \qquad
    \begin{aligned}
       & \Fin[0] \defeq \bot                      \\
       & \Fin[\suc[n]] \defeq \top \sqcup \Fin[n]
    \end{aligned}
  \end{gather*}
  Note that $\Fin[n]$ is a $\hSet$, and we use the definitions interchangeably.
\end{definition}

@@

The groupoid $\UFin$ has two symmetric monoidal structures, the additive and the multiplicative one, and the
multiplicative tensor product distributes over the additive one. To construct these, we first define some equivalences
on $\Fin$, and some general type isomorphisms.
